Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-06 pecomplexity.tex
fixing bibfor revtex
more pre formatting

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%showpacs,preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
% aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}
%\documentclass{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{amssymb}
%\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{amsthm}

@@

\title{Determinism, Complexity, and Predictability in Computer Performance\\
Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
\\ Quantifying Predictability through Structural Complexity \\
{\color{green}Quantifying Predictability through Structural Complexity} 
\\ Quantifying Predictability using Structural Complexity 
\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
}

@@

\author{Joshua Garland}
% \altaffiliation[Also at ]{}%Lines break automatically or can be forced with \\
%\affiliation{%
% Department of Computer Science, University of Colorado at %Boulder. Colorado, USA
%}%
\email{joshua.garland@colorado.edu}
\author{Ryan James}%
 \email{ryan.james@colorado.edu}
\affiliation{%
 Department of Computer Science, University of Colorado at Boulder. Colorado, USA
 %This line break forced with \textbackslash\textbackslash
}%
%\collaboration{MUSO Collaboration}%\noaffiliation
\author{Elizabeth Bradley}\email{lizb@colorado.edu}
\affiliation{
 Department of Computer Science, University of Colorado at Boulder. Colorado, USA
}%
\affiliation{
 Santa Fe Institute, New Mexico, USA\\
 %This line break forced% with \\
}%

@@

\begin{abstract}
[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
  Among other things, that implies that one should be able to use
  deterministic forecast rules to predict aspects of their behavior.
  That statement is sometimes---but not always---true. The memory and
  processor loads of some simple programs are easy to predict, for
  example, but those of more-complex programs like \gcc are not.
%%%%%%%%%%%%%%%%%%%%%%
%% I had to change all the \verb|blah| entries because they caused
%% latex to barf if they were in figure captions.  Odd bug.
%%%%%%%%%%%%%%%%%%%%%%
  The goal of this paper is to determine why that is the case. We
  conjecture that, in practice, complexity can effectively overwhelm
  the predictive power of deterministic forecast models. To explore
  that, we build models of a number of performance traces from
  different programs running on different Intel-based computers. We
  then calculate the \emph{permutation entropy}---a temporal entropy
  metric that uses ordinal analysis---of those traces and correlate
  those values against the prediction success.
\begin{description}
\item[Usage]
Secondary publications and information retrieval purposes.
\item[PACS numbers]
May be entered using the \verb+\pacs{#1}+ command.
\item[Structure]
You may use the \texttt{description} environment to structure your abstract;
use the optional argument of the \verb+\item+ command to give the category of each item. 
\end{description}
\end{abstract}

\pacs{Valid PACS appear here}% PACS, the Physics and Astronomy
                             % Classification Scheme.
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

\bibliographystyle{unsrt}
\bibliography{bibliofile}

>>>>

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%showpacs,preprintnumbers,
nofootinbib,
nobibnotes,
%bibnotes,
 amsmath,amssymb,
% aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}
%\documentclass{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parIEEEabrvskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{amssymb}
%\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{amsthm}

@@

\title{
%Determinism, Complexity, and Predictability in Computer Performance\\
%Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
%\\ 
{\color{green}Quantifying Predictability through Structural Complexity} 
%\\ 
%Quantifying Predictability using Structural Complexity 
\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
}

@@

\author{Joshua Garland}\email{joshua.garland@colorado.edu}
\author{Ryan James}%
 \email{ryan.james@colorado.edu}
\affiliation{%
 Department of Computer Science, University of Colorado at Boulder. Colorado, USA}%
%\collaboration{MUSO Collaboration}%\noaffiliation
\author{Elizabeth Bradley}\email{lizb@colorado.edu}
\affiliation{
 Department of Computer Science, University of Colorado at Boulder. Colorado, USA
}%
\affiliation{
 Santa Fe Institute, New Mexico, USA\\
 %This line break forced% with \\
}%

@@

\begin{abstract}
{\color{green}From the site:It should be about 5\% of the length of the article, but less than 500 words.}[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
  Among other things, that implies that one should be able to use
  deterministic forecast rules to predict aspects of their behavior.
  That statement is sometimes---but not always---true. The memory and
  processor loads of some simple programs are easy to predict, for
  example, but those of more-complex programs like \gcc are not.
%%%%%%%%%%%%%%%%%%%%%%
%% I had to change all the \verb|blah| entries because they caused
%% latex to barf if they were in figure captions.  Odd bug.
%%%%%%%%%%%%%%%%%%%%%%
  The goal of this paper is to determine why that is the case. We
  conjecture that, in practice, complexity can effectively overwhelm
  the predictive power of deterministic forecast models. To explore
  that, we build models of a number of performance traces from
  different programs running on different Intel-based computers. We
  then calculate the \emph{permutation entropy}---a temporal entropy
  metric that uses ordinal analysis---of those traces and correlate
  those values against the prediction success.
\begin{description}
\item[Usage]
Secondary publications and information retrieval purposes.
\item[PACS numbers]
May be entered using the \verb+\pacs{#1}+ \pacs{05.45.Tp} command.
Time series analysis (nonlinear dynamics) 05.45.Tp
Information theory, 89.70.-a
	channel capacity in, 89.70.Kn
	communication complexity in, 89.70.Hj
	computational complexity in, 89.70.Eg
	entropy in, 89.70.Cf
	general biological information, 87.10.Vg
	in neuroscience, 87.19.lo

Nonlinear dynamics, 05.45.-a

	89.75.Kd	Patterns (complex systems)

Computational Physics	
95.75.Wx	Time series analysis, time variability
95.75.Pq	Mathematical procedures and computer techniques
07.05.-t	Computers in experimental physics

\item[Structure]
You may use the \texttt{description} environment to structure your abstract;
use the optional argument of the \verb+\item+ command to give the category of each item. 
\end{description}
\end{abstract}

\pacs{Valid PACS appear here 95.75.Wx	}% PACS, the Physics and Astronomy
                             % Classification Scheme.
%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle

@@

\bibliographystyle{aipauth4-1}
\bibliography{bibliofile}

### 2014-03-06 results.tex
putting it in Phys Rev E Format
more pre formatting

 \section{Predictability, Complexity, and Permutation Entropy {\color{blue}  EDITABLE}} % (fold)
 \label{sec:results}
% I changed the title of this section because there are lots of
% results in the modeling section too.  The results HERE are about the
% relationship between predictability and WPE

@@

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=0.6\textwidth]{figs/LMA_prediction_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm
  0.282$, and $20.870 \pm 0.192$ respectively.  {\color{red}Replace
    the top plot with a plot of the naive results---but running from 0
    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
  dashed line is the same as in the previous figure [[Ryan, is this
      correct?]].  Numerical values, including means and standard
  deviations of the errors, can be found in Table~\ref{tab:error} in
  the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of these three plots.)  The WPE
values do not vary very much across trials.  For most traces, the
variance in MASE scores is low as well, resulting in small, tight
clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
instance---the MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots: clusters of predictions generated
with the nonlinear LMA method are closer to the dashed line, whereas
the naive and ARIMA prediction clusters are more spread out [[need to
    revisit this wording when get the third image]].  This is a
geometric validation of the second conjecture in this paper, as
described in the following paragraphs.

@@

\svdone behaves very differently than the other seven systems in this
study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
prediction methods do quite well, yielding MASE scores of $0.82 \pm
0.08$ (LMA) and $0.714 \pm 0.07$ (ARIMA).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven clusters.  Also, the MASE scores of the predictions of
this system produced by the \naive method are highly inconsistent:
$2.67 \pm 4.33$.  These effects, we believe, are artifacts of the way
MASE is calculated.  Recall that MASE scores are scaled relative to a
random-walk forecast, which uses the current value as the prediction.
This strategy works very badly on signals with frequent, large, rapid
transitions.  Consider a signal that oscillates from one end of its
range to the other at every step.  A signal like this will have a low
WPE, much like {\tt col\_major}.  However, a random-walk forecast of
this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the MASE score, this effect can shift points leftwards on a WPE
vs. MASE plot, and that is exactly why the \svdone clusters for ARIMA
and LMA are above the dashed line in Figure~\ref{fig:wpe_vs_mase_all}.
This time series, which is shown in closeup in
Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed line
in Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

 \section{Predictability, Complexity, and Permutation Entropy %{\color{blue}  EDITABLE}
 } % (fold)
 \label{sec:results}
% I changed the title of this section because there are lots of
% results in the modeling section too.  The results HERE are about the
% relationship between predictability and WPE

@@

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:wwpe}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.  Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of
these 45 predictions for each system: i.e., the lowest error over all
15 trials and all three methods for each of the eight programs.  The
WPE is plotted against the corresponding MASE value here in order to
bring out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.  [[Ryan, is that correct?]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
\caption{WPE vs. MASE for all trials, methods, and systems.  \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons; their MASE scores are $2.676 \pm 4.328$, $31.386 \pm
  0.282$, and $20.870 \pm 0.192$ respectively.  {\color{red}Replace
    the top plot with a plot of the naive results---but running from 0
    to 3.1 on the horizontal axis and excluding SVD 1, 3, and 5}.  The
  dashed line is the same as in the previous figure [[Ryan, is this
      correct?]].  Numerical values, including means and standard
  deviations of the errors, can be found in Table~\ref{tab:error} in
  the appendix.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of these three plots.)  The WPE
values do not vary very much across trials.  For most traces, the
variance in MASE scores is low as well, resulting in small, tight
clusters.  In some cases---ARIMA predictions of {\tt col\_major}, for
instance---the MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots: clusters of predictions generated
with the nonlinear LMA method are closer to the dashed line, whereas
the naive and ARIMA prediction clusters are more spread out [[need to
    revisit this wording when get the third image]].  This is a
geometric validation of the second conjecture in this paper, as
described in the following paragraphs.

@@

\svdone behaves very differently than the other seven systems in this
study: though its WPE is very high ($0.95 \pm 0.02$), two of the three
prediction methods do quite well, yielding MASE scores of $0.82 \pm
0.08$ (LMA) and $0.714 \pm 0.07$ (ARIMA).  This pushes the
corresponding clusters of points in the bottom two images in
Figure~\ref{fig:wpe_vs_mase_all} well above the trend followed by the
other seven clusters.  Also, the MASE scores of the predictions of
this system produced by the \naive method are highly inconsistent:
$2.67 \pm 4.33$.  These effects, we believe, are artifacts of the way
MASE is calculated.  Recall that MASE scores are scaled relative to a
random-walk forecast, which uses the current value as the prediction.
This strategy works very badly on signals with frequent, large, rapid
transitions.  Consider a signal that oscillates from one end of its
range to the other at every step.  A signal like this will have a low
WPE, much like {\tt col\_major}.  However, a random-walk forecast of
this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the MASE score, this effect can shift points leftwards on a WPE
vs. MASE plot, and that is exactly why the \svdone clusters for ARIMA
and LMA are above the dashed line in Figure~\ref{fig:wpe_vs_mase_all}.
This time series, which is shown in closeup in
Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed line
in Figure~\ref{fig:wpe_vs_mase_best}.
